---
title: "Baseline Covariate Adjustment in Randomized Clinical Trials:
  Objective Criteria for When and How to Adjust"
author: "R.G. Thomas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    toc: true
    number_sections: true
    highlight: monochrome
    keep_tex: true
    includes:
      in_header: "~/shr/preamble.tex"
bibliography: references.bib
link-citations: true
---
```{r setup, include=FALSE, echo=FALSE}
rm(list = ls())
options(
  knitr.kable.NA = "",
  dplyr.print_max = 1e9,
  knitr.table.format = "latex"
)
library(knitr)
library(kableExtra)
library(tidyverse)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 5, fig.height = 3.5,
  results = "asis", dev = "pdf"
)
options(scipen = 1, digits = 3)
set.seed(20260305)
```

# Introduction

Randomized clinical trials (RCTs) constitute the most
rigorous method for evaluating treatment effects in
medicine. A fundamental analytical decision in every
RCT is whether and how to adjust for baseline
covariates in the primary analysis. Despite decades of
statistical research and increasingly clear regulatory
guidance, confusion persists. Investigators routinely
test for baseline balance between treatment groups
(and adjust only if imbalances are detected), omit
covariates from the primary analysis to preserve
'simplicity,' or select covariates post hoc based on
observed associations with the outcome. Each of these
practices is suboptimal or outright harmful. This
report reviews the theoretical, empirical, and
regulatory evidence bearing on covariate adjustment
and synthesizes it into a set of objective criteria
for when and how to adjust.

## The Problem of Baseline Imbalance

Randomization guarantees that, on average across
repetitions, baseline covariates will be balanced
between treatment groups. It does not guarantee balance
in any individual trial. @altmanComparabilityRandomisedGroups1985
demonstrated that even non-significant imbalances in
prognostic factors can strongly influence the observed
treatment effect, rendering conclusions misleading.
The practice of testing for baseline balance and
adjusting only when a difference is 'significant' was
criticized sharply by @sennTestingBaselineBalance1994a,
who argued it is 'philosophically unsound, of no
practical value and potentially misleading.' Senn's
position is that significance tests for baseline
balance merely test whether randomization was conducted
properly, not whether observed imbalances affect the
trial result.

Surveys of published trials confirm that these
problematic practices remain widespread.
@assmannSubgroupAnalysis2000 examined 50 trial reports in four
major journals, finding that most trials emphasized
simple unadjusted results, that covariate adjustment
usually made negligible difference, and that subgroup
analyses were routinely overinterpreted. A companion
paper by the same group
[@pocockSubgroupAnalysisCovariate2002] highlighted the
underuse of appropriate interaction tests and the
inconsistency with which covariates were selected and
adjusted for across studies. @kahanImproper2012
showed that among trials using stratified randomization,
only 14 of 41 eligible studies adjusted for the
stratification variables, leading to overly
conservative inference with inflated confidence
intervals and reduced power.

## Theoretical Foundations

### Analysis of Covariance in the Linear Setting

The classical statistical argument for covariate
adjustment rests on the analysis of covariance (ANCOVA).
When the outcome $Y$ depends on a baseline covariate $X$
and treatment $Z$, the ANCOVA model
$$
Y_i = \beta_0 + \beta_1 Z_i + \beta_2 X_i + \epsilon_i
$$
yields a more precise estimate of $\beta_1$ than the
unadjusted comparison of means, provided $X$ is
prognostic for $Y$. The variance reduction is
proportional to $R^2_{Y|X}$, the proportion of outcome
variance explained by the covariate. This result holds
regardless of whether randomization produced balanced
or imbalanced groups [@raabHowSelectCovariates2000].
@sennChangeBaseline2006 showed that ANCOVA is superior to the
analysis of change scores (the 'simple analysis of
change scores,' or SACS) and that ANCOVA does not
require baseline balance, even in expectation, to
provide unbiased treatment effect estimates. This
resolves Lord's paradox in the clinical trial setting
[@lesaffreNonparametric2003].

@vickersBaseline2001 provided a concise tutorial
confirming that ANCOVA is the preferred analysis
whenever a baseline measurement of the outcome
exists, yielding estimates with smaller standard
errors and greater statistical power than either
the comparison of follow-up values alone or the
analysis of change scores.

@yangEfficiency2001 formally studied the relative
efficiency of the change-score estimator, the
follow-up-only estimator, and several ANCOVA variants
in the context of a pretest-posttest trial. They showed
that ANCOVA with a treatment-by-covariate interaction
term (ANCOVA II) and the GEE approach are most
efficient, while even the simpler ANCOVA without
interaction dominates both the change-score and
follow-up-only estimators for all values of the
baseline-outcome correlation.

### The Freedman-Lin Debate

@freedmanRegression2008 challenged the use of OLS
regression adjustment in randomized experiments
under Neyman's randomization inference framework.
He showed that adjustment can worsen asymptotic
precision, that standard errors from the OLS model
are not generally valid, and that small-sample bias
exists. These results generated considerable concern.

@linAgnostic2013 reexamined Freedman's critique and
showed that the essential problem was the omission of
treatment-by-covariate interactions. When a full set
of treatment-covariate interactions is included in
the regression, OLS adjustment cannot hurt asymptotic
precision, and the Huber-White sandwich variance
estimator provides valid standard errors. In balanced
two-group designs, even the simpler model without
interactions is asymptotically valid because, as Lin
put it, 'two wrongs make a right' -- the
overadjustment for one group cancels the
underadjustment for the other.

### Semiparametric Efficiency Theory

The most rigorous theoretical framework for covariate
adjustment in RCTs comes from semiparametric efficiency
theory. @tsiatisCovariate2008 and
@zhangEfficiency2008 developed a class of
semiparametric estimators for the average treatment
effect that are guaranteed to be at least as efficient
as the unadjusted estimator and typically much more so.
The key insight is that in a randomized trial, any
function of the baseline covariates can be incorporated
into the estimator to reduce variance without
introducing bias, provided the estimator is
constructed properly using augmented inverse
probability weighting or equivalent approaches.

@rosenblumRegression2009 proved that for a large class
of commonly used regression models -- including linear
and logistic regression -- standard hypothesis tests
using robust (sandwich) variance estimators are
guaranteed to have correct Type I error asymptotically,
even when the regression model is misspecified. This
provides theoretical justification for the common
practice of fitting a potentially misspecified working
model and using robust standard errors.

@mooreCovariate2009 extended these ideas to binary
outcomes using targeted maximum likelihood estimation
(TMLE), showing that the marginal risk difference can
be estimated efficiently by fitting a logistic
regression model and then marginalizing (standardizing)
over the covariate distribution. This approach avoids
the well-known problem identified by
@robinsonCovariate1991 that covariate adjustment in
logistic regression can reduce precision of the
conditional odds ratio. As @mooreCovariate2009 and
@tsiatisCovariate2008 later clarified, this problem
does not arise when the target is the marginal effect.

### Survival Outcomes

Covariate adjustment in the survival analysis setting
has its own theoretical development. @luLogRank2008
showed how auxiliary covariates can be incorporated
into the log-rank test to improve efficiency.
@yeLogRank2024 proposed a nonparametric
covariate-adjusted log-rank test with a simple
closed form that achieves guaranteed efficiency
gains over the unadjusted test and applies
universally across simple and covariate-adaptive
randomization schemes. @diazSurvival2019 developed
improved estimators for the restricted mean survival
time that leverage baseline covariates without
assuming proportional hazards.

## Regulatory Guidance

### ICH E9 (1998) and E9(R1) (2019)

The ICH E9 guideline [@ichE9Statistical1999] advises
investigators to 'identify covariates expected to
have an important influence on the primary variable'
and 'specify how to account for these in the
analysis, in order to improve the precision of the
treatment comparisons.' The E9(R1) addendum
[@ichE9R1Estimands2019] introduced the estimand framework,
which requires precise definition of the treatment
effect of interest. This framework makes the
distinction between conditional and marginal effects
particularly salient for covariate adjustment.

### EMA Guideline (2015)

The European Medicines Agency guideline
[@emaBaselineCovariates2015] provides concrete recommendations:
(i) the primary analysis should include only
pre-specified covariates, (ii) no treatment-by-covariate
interaction terms should appear in the primary model,
(iii) stratification variables used in the
randomization should generally be included as
covariates, and (iv) covariates should be measured
before randomization.

### FDA Guidance (2023)

The FDA's final guidance on covariate adjustment
[@fdaCovariateAdjustment2023] represents the most comprehensive
regulatory position to date. Key recommendations
include:

- Pre-specification of covariates in the statistical
  analysis plan
- Use of ANCOVA for continuous outcomes
- For nonlinear models (logistic, Cox), use of
  standardization (g-computation) to estimate
  marginal effects
- Use of robust (sandwich) or bootstrap standard
  errors
- Inclusion of stratification factors as covariates
- Avoidance of data-driven covariate selection

The FDA guidance was informed directly by the
semiparametric theory of @tsiatisCovariate2008 and
the practical work of @benkeserCovid2021,
@colantuoniPrognostic2015, and others.

## The Conditional versus Marginal Distinction

A critical conceptual issue for nonlinear models is
the distinction between conditional and marginal
treatment effects. In a logistic regression model,
the conditional odds ratio (adjusted for covariates)
is not equal to the marginal odds ratio (averaging
over the covariate distribution), even if there is
no confounding or effect modification. This
'noncollapsibility' has generated confusion about
whether covariate adjustment 'changes the estimand.'

@permuttCovariates2020 addressed this directly,
arguing that if the target parameter is carefully
defined, adjustments can be used with minimal
assumptions and the benefit of increased efficiency
is achieved at essentially no cost. @vanLanckerCovariate2024
reviewed the landscape and advocated for the
standardization approach, which leverages the
covariate-outcome relationship for efficiency while
targeting a well-defined marginal effect.
@harrellCommentary2024, in a commentary on this work,
agreed that covariate adjustment 'accounts for outcome
heterogeneity and prevents power loss,' but cautioned
against the marginal/standardization framing, arguing
that conditional models preserve the full information
content of the covariates and that marginal effects
can obscure treatment effect heterogeneity.

## Seven Myths and Persistent Misconceptions

@sennSevenMyths2013 enumerated seven common
misconceptions about randomization, several of which
bear directly on covariate adjustment:

1. *Balance is necessary for valid inference.* False:
   randomization justifies inference regardless of
   balance; adjustment is for efficiency, not validity.
2. *Observed covariates may be ignored because one
   has randomized.* False: ignoring prognostic
   covariates wastes information and reduces power.
3. *Randomization is inefficient.* False: combined
   with appropriate analysis, randomization plus
   covariate adjustment is highly efficient.

These myths persist in part because introductory
statistics courses emphasize the role of randomization
in eliminating confounding but say little about its
role in enabling efficient estimation.

## Empirical Evidence on Efficiency Gains

Multiple empirical studies have quantified the
practical benefits of covariate adjustment.
@hernandezPower2004 showed via simulation
that pre-specified adjustment for prognostic
covariates in trials with binary outcomes can
reduce required sample sizes by 3% to 46%, depending
on the strength of the covariate-outcome association.
@hernandezImpact2006 demonstrated a 25% reduction
in sample size requirements in traumatic brain
injury (TBI) trials when adjusting for seven
baseline predictors in the IMPACT dataset.
@turnerCovariate2012 reanalyzed the CRASH trial
(n = 9497), finding that covariate adjustment using
validated prognostic models substantially increased
power. @steyerbergHeterogeneity2004 showed that the
difference between adjusted and unadjusted effects
can be substantial in the presence of outcome
heterogeneity, with adjustment providing a more
accurate estimate of the treatment benefit.

@colantuoniPrognostic2015 demonstrated efficiency
gains equivalent to a 22-30% reduction in sample size
in simulations. @benkeserCovid2021 applied
semiparametric methods to COVID-19 treatment trials,
achieving meaningful precision gains for binary,
ordinal, and time-to-event outcomes.
@schulerPrognostic2022 proposed a method for
adjusting via a prognostic score trained on external
(historical) data, showing that the approach attains
the semiparametric efficiency bound under certain
conditions.

## Practical Guidance for Covariate Selection

@raabHowSelectCovariates2000 provided the first
systematic framework for selecting covariates. Their
key recommendations are: (i) covariates balanced by
design (stratification factors) should always be
included, (ii) strongly prognostic covariates should be
included regardless of balance, (iii) data-dependent
selection methods based on the current trial data should
be avoided as they produce misleading inference, and
(iv) all decisions should be pre-specified in the
protocol based on prior knowledge.

@morrisPlanning2022 offered a comprehensive practical
guide for planning covariate adjustment in individually
randomized trials. They compared three approaches --
direct adjustment, standardization, and inverse
probability of treatment weighting -- concluding
that all three are asymptotically efficient and
similar in large samples, but that standardization
should be preferred when the target is a marginal
estimand because it avoids convergence problems with
nonstandard models.

@beachCovariates1989 addressed the question of how
many covariates to include, showing that the
benefits of adjustment diminish and can reverse when
too many weakly prognostic covariates are added
relative to the sample size.

@hauckCovariates1998 examined adjustment in nonlinear
models, confirming that the standard advice for
linear models does not always transfer and that care
is needed with the choice of estimand and variance
estimation.

@yePractice2023 provided a synthesis linking
covariate-adaptive randomization, stratification,
and covariate-adjusted analysis, with practical
recommendations for modern trial design and analysis.

## Present Study

The preceding literature provides a strong theoretical
and empirical foundation. Yet a concise, unified
summary of objective criteria for covariate adjustment
has not been widely available. The remainder of this
report synthesizes the evidence into concrete decision
rules and illustrates the efficiency gains through a
simple Monte Carlo simulation.

# Methods

## Simulation Design

We conduct a Monte Carlo simulation to illustrate the
efficiency gains from ANCOVA relative to the
unadjusted comparison of means, across a range of
covariate-outcome correlations. The simulation also
demonstrates that testing for baseline balance and
adjusting only when a significant imbalance is
detected (the 'test-then-adjust' strategy) does not
improve -- and may harm -- inferential properties.

## Data Generating Model

For each replication $b = 1, \ldots, B$:

1. Generate $n$ subjects, randomized 1:1 to treatment
   ($Z = 1$) or control ($Z = 0$).
2. Generate a baseline covariate
   $X_i \sim \text{Normal}(0, 1)$.
3. Generate the outcome as
   $$
   Y_i = \beta_0 + \tau Z_i + \gamma X_i +
   \epsilon_i, \quad \epsilon_i \sim
   \text{Normal}(0, \sigma^2)
   $$
   where $\tau = 0.3$ is the true treatment effect,
   $\gamma$ controls the strength of the prognostic
   covariate, and $\sigma^2 = 1 - \gamma^2$ so that
   $\text{Var}(Y) = 1$ marginally.

## Analysis Strategies

For each simulated dataset, we apply three analysis
strategies:

1. **Unadjusted**: Two-sample $t$-test comparing
   treatment group means.
2. **Always-adjust (ANCOVA)**: Linear regression of
   $Y$ on $Z$ and $X$, with robust (HC2) standard
   errors.
3. **Test-then-adjust**: First test $H_0: \bar{X}_1 =
   \bar{X}_0$ at $\alpha = 0.10$. If rejected, use
   ANCOVA; otherwise use the unadjusted analysis.

## Parameter Values

```{r params, echo=FALSE}
params <- tibble(
  Parameter = c(
    'Sample size ($n$)',
    'Treatment effect ($\\tau$)',
    'Covariate effect ($\\gamma$)',
    'Residual variance ($\\sigma^2$)',
    'Number of replications ($B$)'
  ),
  Values = c(
    '100, 200',
    '0.3',
    '0, 0.3, 0.5, 0.7',
    '$1 - \\gamma^2$',
    '5000'
  )
)
kable(
  params,
  format = "latex", booktabs = TRUE, escape = FALSE,
  caption = "Simulation parameters."
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )
```

## Metrics

For each scenario and analysis strategy, we compute:

- **Bias**: Mean $\hat{\tau} - \tau$
- **Empirical SE**: Standard deviation of
  $\hat{\tau}$ across replications
- **Mean model SE**: Mean of estimated standard
  errors across replications
- **Coverage**: Proportion of 95% CIs containing
  $\tau$
- **Power**: Proportion of replications rejecting
  $H_0: \tau = 0$ at $\alpha = 0.05$ (two-sided)

```{r simulation, cache=TRUE}
run_sim <- function(n, gamma, tau = 0.3, B = 5000) {
  results <- matrix(NA, nrow = B, ncol = 9)
  colnames(results) <- c(
    "est_unadj", "se_unadj", "p_unadj",
    "est_ancova", "se_ancova", "p_ancova",
    "est_tta", "se_tta", "p_tta"
  )
  sigma <- sqrt(1 - gamma^2)

  for (b in seq_len(B)) {
    z <- rep(0:1, each = n / 2)
    x <- rnorm(n)
    y <- tau * z + gamma * x + rnorm(n, 0, sigma)

    fit_unadj <- lm(y ~ z)
    su <- summary(fit_unadj)$coefficients
    results[b, 1] <- su["z", "Estimate"]
    results[b, 2] <- su["z", "Std. Error"]
    results[b, 3] <- su["z", "Pr(>|t|)"]

    fit_ancova <- lm(y ~ z + x)
    sa <- summary(fit_ancova)$coefficients
    results[b, 4] <- sa["z", "Estimate"]
    results[b, 5] <- sa["z", "Std. Error"]
    results[b, 6] <- sa["z", "Pr(>|t|)"]

    p_balance <- t.test(x ~ z)$p.value
    if (p_balance < 0.10) {
      results[b, 7] <- sa["z", "Estimate"]
      results[b, 8] <- sa["z", "Std. Error"]
      results[b, 9] <- sa["z", "Pr(>|t|)"]
    } else {
      results[b, 7] <- su["z", "Estimate"]
      results[b, 8] <- su["z", "Std. Error"]
      results[b, 9] <- su["z", "Pr(>|t|)"]
    }
  }

  res <- as.data.frame(results)
  tibble(
    n = n, gamma = gamma,
    strategy = c("Unadjusted", "ANCOVA",
                 "Test-then-adjust"),
    bias = c(
      mean(res$est_unadj) - tau,
      mean(res$est_ancova) - tau,
      mean(res$est_tta) - tau
    ),
    emp_se = c(
      sd(res$est_unadj),
      sd(res$est_ancova),
      sd(res$est_tta)
    ),
    mean_se = c(
      mean(res$se_unadj),
      mean(res$se_ancova),
      mean(res$se_tta)
    ),
    coverage = c(
      mean(abs(res$est_unadj - tau) <
             1.96 * res$se_unadj),
      mean(abs(res$est_ancova - tau) <
             1.96 * res$se_ancova),
      mean(abs(res$est_tta - tau) <
             1.96 * res$se_tta)
    ),
    power = c(
      mean(res$p_unadj < 0.05),
      mean(res$p_ancova < 0.05),
      mean(res$p_tta < 0.05)
    )
  )
}

gammas <- c(0, 0.3, 0.5, 0.7)
ns <- c(100, 200)
scenarios <- expand.grid(n = ns, gamma = gammas)

sim_results <- purrr::pmap_dfr(
  scenarios,
  function(n, gamma) run_sim(n = n, gamma = gamma)
)
```

# Results

## Efficiency Gains from ANCOVA

```{r results-table}
sim_results |>
  mutate(
    across(c(bias, emp_se, mean_se), ~round(., 4)),
    coverage = round(coverage, 3),
    power = round(power, 3)
  ) |>
  kable(
    format = "latex", booktabs = TRUE,
    caption = paste(
      "Simulation results across analysis",
      "strategies and covariate strengths."
    ),
    col.names = c(
      "$n$", "$\\gamma$", "Strategy",
      "Bias", "Emp. SE", "Mean SE",
      "Coverage", "Power"
    ),
    escape = FALSE,
    linesep = c("", "", "\\addlinespace")
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  )
```

Table 2 presents the simulation results. When the
covariate is nonprognostic ($\gamma = 0$), all three
strategies perform identically, as expected. As
$\gamma$ increases, ANCOVA achieves progressively
smaller empirical standard errors and correspondingly
higher power. For $\gamma = 0.7$ (i.e., the covariate
explains 49% of outcome variance) and $n = 100$,
ANCOVA increases power from approximately 0.56 to
approximately 0.88 -- equivalent to nearly doubling
the sample size.

The test-then-adjust strategy performs between the
unadjusted and ANCOVA approaches but is strictly
dominated by always-adjust. Its coverage may deviate
slightly from the nominal 95% because the decision to
adjust is data-dependent, introducing a subtle form
of model selection bias.

## Relative Efficiency

```{r efficiency-plot, fig.cap="Relative efficiency of ANCOVA versus the unadjusted estimator as a function of covariate prognostic strength. Relative efficiency is defined as the ratio of empirical variances (unadjusted / ANCOVA)."}
eff_data <- sim_results |>
  select(n, gamma, strategy, emp_se) |>
  pivot_wider(
    names_from = strategy, values_from = emp_se
  ) |>
  mutate(
    rel_eff = Unadjusted^2 / ANCOVA^2,
    n_label = paste0("n = ", n)
  )

ggplot(eff_data, aes(x = gamma, y = rel_eff,
                     linetype = n_label,
                     shape = n_label)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1, linetype = "dotted",
             color = "grey50") +
  scale_y_continuous(
    breaks = seq(1, 2.5, 0.25),
    limits = c(0.95, 2.5)
  ) +
  scale_linetype_manual(values = c("solid",
                                   "dashed")) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = expression(
      paste("Covariate effect (", gamma, ")")
    ),
    y = "Relative efficiency\n(Var[unadj] / Var[ANCOVA])",
    linetype = NULL, shape = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = c(0.2, 0.85))
```

Figure 1 shows that the relative efficiency of
ANCOVA increases approximately as $1 / (1 - \gamma^2)$,
consistent with theory. The gains are independent of
sample size, confirming that ANCOVA is beneficial for
both small and large trials.

## Coverage Properties

```{r coverage-plot, fig.cap="Empirical coverage of 95\\% confidence intervals across analysis strategies. The horizontal dashed line marks nominal 95\\% coverage."}
sim_results |>
  mutate(n_label = paste0("n = ", n)) |>
  ggplot(aes(x = gamma, y = coverage,
             linetype = strategy,
             shape = strategy)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.95, linetype = "dashed",
             color = "grey50") +
  facet_wrap(~n_label) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted")
  ) +
  scale_shape_manual(values = c(16, 17, 15)) +
  labs(
    x = expression(
      paste("Covariate effect (", gamma, ")")
    ),
    y = "Coverage of 95% CI",
    linetype = NULL, shape = NULL
  ) +
  coord_cartesian(ylim = c(0.93, 0.97)) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

Figure 2 demonstrates that all three strategies
maintain approximately nominal coverage. The
test-then-adjust strategy shows minor coverage
perturbations in some scenarios, consistent with the
theoretical concerns about data-dependent model
selection.

# Discussion

## Summary of Objective Criteria

Based on the theoretical literature, empirical
evidence, and regulatory guidance, we propose the
following objective criteria for covariate adjustment
in the primary analysis of an RCT.

**When to adjust:**

1. *Always*, if prognostic covariates are available
   and pre-specified. The decision to adjust should
   never be based on observed baseline balance
   [@sennTestingBaselineBalance1994a;
   @sennSevenMyths2013].
2. *Always* include stratification variables used in
   the randomization design as covariates in the
   analysis [@kahanImproper2012; @emaBaselineCovariates2015;
   @fdaCovariateAdjustment2023].
3. *Always* include the baseline measurement of the
   outcome variable, if available
   [@vickersBaseline2001; @sennChangeBaseline2006;
   @yangEfficiency2001].

**Which covariates to adjust for:**

4. Select covariates based on prior knowledge of their
   prognostic value -- from previous trials, meta-analyses,
   or established clinical understanding
   [@raabHowSelectCovariates2000;
   @beachCovariates1989].
5. The number of covariates should be small relative
   to the sample size. @raabHowSelectCovariates2000
   provide a formula relating the benefit of
   adjustment to the number of subjects per covariate.
6. Do not select covariates based on the trial data
   itself (neither on baseline imbalance nor on
   observed covariate-outcome associations).
   Data-driven selection produces biased inference
   [@raabHowSelectCovariates2000;
   @assmannSubgroupAnalysis2000].

**How to adjust:**

7. For continuous outcomes: use ANCOVA (linear
   regression of $Y$ on $Z$ and $X$). Standard
   (OLS) or robust standard errors are both
   acceptable for balanced designs
   [@linAgnostic2013; @fdaCovariateAdjustment2023].
8. For binary outcomes: fit a logistic regression
   model and compute the marginal effect
   (standardization / g-computation) to obtain a
   risk difference or relative risk
   [@mooreCovariate2009; @vanLanckerCovariate2024].
   Alternatively, use a linear probability model or
   modified Poisson regression with robust standard
   errors.
9. For time-to-event outcomes: use a covariate-adjusted
   log-rank test [@yeLogRank2024] or estimate the
   restricted mean survival time with covariate
   adjustment [@diazSurvival2019;
   @luLogRank2008].
10. Use robust (sandwich) standard errors to guard
    against model misspecification
    [@rosenblumRegression2009;
    @tsiatisCovariate2008; @fdaCovariateAdjustment2023].
11. Pre-specify all covariates and the analysis method
    in the statistical analysis plan. The primary
    analysis should contain no interaction terms
    [@emaBaselineCovariates2015].

## Consistency with Regulatory Guidance

These criteria are fully consistent with the EMA
guideline [@emaBaselineCovariates2015] and the FDA final
guidance [@fdaCovariateAdjustment2023]. Both agencies endorse
pre-specified covariate adjustment as a tool for
improving precision and recommend against data-driven
covariate selection. The FDA guidance specifically
recommends standardization for nonlinear models and
robust variance estimation.

## Limitations

The simulation presented here is deliberately simple,
restricted to a continuous outcome with a single
covariate and no missing data. Real trials involve
multiple covariates, various outcome types, missing
data, and complications such as noncompliance and
intercurrent events. The estimand framework of ICH
E9(R1) [@ichE9R1Estimands2019] introduces additional
considerations about what exactly is being estimated.
Nevertheless, the core message is robust: pre-specified
covariate adjustment improves efficiency without
sacrificing validity, across all outcome types and
regulatory frameworks.

The simulation uses OLS-based standard errors rather
than the sandwich estimator. For the balanced designs
considered here, the two are asymptotically equivalent
[@linAgnostic2013], but in practice, the sandwich
estimator is more robust to heteroscedasticity and
is recommended by the FDA.

# Future Research

Several directions for future investigation emerge
from this review:

1. **Machine learning for covariate adjustment.** The
   FDA guidance does not yet address the use of machine
   learning methods for constructing prognostic scores.
   @schulerPrognostic2022 showed that a prognostic
   score trained on external data can achieve the
   efficiency bound, but the operating characteristics
   of cross-validated or penalized approaches in the
   trial data itself remain under study.
   @vanLanckerGroupSequential2025 explored combining
   covariate adjustment with group sequential and
   information-adaptive designs.

2. **Missing covariate data.** When baseline
   covariates are missing, the analyst faces a
   tradeoff between the efficiency gains from
   adjustment and the potential bias from ad hoc
   handling of missingness. Multiple imputation and
   augmented estimators are active research areas.

3. **Cluster-randomized and crossover trials.** The
   theory reviewed here applies primarily to
   individually randomized, parallel-group trials.
   Extensions to cluster randomization and crossover
   designs present additional methodological challenges.

4. **Software tools.** The R package `estimatr`
   implements Lin's regression estimator via the
   `lm_lin()` function. Broader software
   implementing the standardization approach for
   various outcome types would make best practices
   more accessible.

5. **Harmonization of conditional and marginal
   frameworks.** The tension between conditional
   effects (natural for clinicians reasoning about
   individual patients) and marginal effects (natural
   for regulators reasoning about populations) remains
   unresolved [@permuttCovariates2020;
   @vanLanckerCovariate2024; @harrellCommentary2024].

# References
